---
title: "Supplemental Iron Intake by Age and Gender"
subtitle: "NHANES 2021-2023"
author: "Alissa Weng"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## About this project

Originally completed as a four-person group project for PH 142
(Introduction to Probability and Statistics in Biology and Public
Health) at UC Berkeley, Spring 2026.

My contributions were the data import, visualizations, the
distribution and probability work, the hypotheses, and the Wilcoxon
rank-sum tests.

Ally worked alongside me on the coding and statistical questions;
Bharathi and Eve led the explanatory and interpretive sections. Many
questions were collaborative, and all four of us reviewed and
approved the final version.

## Question

Do older adults report lower average daily iron intake from dietary supplements
over the past 30 days than younger adults, and does this age pattern differ by
gender?

In PPDAC terms, stratifying by gender makes this a comparative relationship
question. Without the gender stratification it would be a descriptive
comparative question.

## Data

Publicly available data from the U.S. National Health and Nutrition Examination
Survey (NHANES), August 2021 to August 2023 cycle. Two files are used:

- `DEMO_L.xpt`, demographics (age, gender)
- `DSQTOT_L.xpt`, total dietary supplements, giving average daily nutrient
  intake from supplements and non-prescription antacids over the past 30 days

Both are downloadable from <https://wwwn.cdc.gov/nchs/nhanes/>.

The sampling frame is the U.S. civilian non-institutionalized population
sampled between August 2021 and August 2023. NHANES uses probability sampling,
so findings generalize to that population over that period.

## Load and merge data

```{r load-merge}
library(foreign)
library(dplyr)

Total_Dietary_Supplements <- read.xport("DSQTOT_L.xpt")
Demographics <- read.xport("DEMO_L.xpt")

write.csv(Total_Dietary_Supplements, "Total_Dietary_Supplements.csv", row.names = FALSE)
write.csv(Demographics, "Demographics.csv", row.names = FALSE)

Total_Dietary_Supplements <- read.csv("Total_Dietary_Supplements.csv")
Demographics <- read.csv("Demographics.csv")

# SEQN is the respondent ID shared by both files.
demo_keep <- Demographics %>% select(SEQN, RIDAGEYR, RIAGENDR)
iron_keep <- Total_Dietary_Supplements %>% select(SEQN, DSQTIRON)

combined <- iron_keep %>% left_join(demo_keep, by = "SEQN")
```

```{r inspect}
dim(combined)
names(combined)
head(combined, 6)
```

## Clean and transform

```{r transform}
# RIAGENDR is coded 1 = Male, 2 = Female.
combined_gender <- combined %>%
  mutate(Gender = factor(RIAGENDR, levels = c(1, 2), labels = c("Male", "Female")))

# Raw iron intake has many zero and near-zero values plus a few large ones,
# so most points pile up at the bottom. A natural log spreads the low end and
# compresses the high outliers.
combined_gender_log <- combined_gender %>%
  mutate(iron_log = log(DSQTIRON))
```

## Linear fits by gender

```{r models}
library(broom)

females <- combined_gender_log %>% filter(Gender == "Female")
males   <- combined_gender_log %>% filter(Gender == "Male")

female_lm <- lm(iron_log ~ RIDAGEYR, data = females)
male_lm   <- lm(iron_log ~ RIDAGEYR, data = males)
total_trend <- lm(DSQTIRON ~ RIDAGEYR, data = combined)

glance(male_lm)   %>% pull(r.squared)
glance(female_lm) %>% pull(r.squared)
glance(total_trend) %>% pull(r.squared)
```

## Plot 1: log intake against age, by gender

```{r scatter, fig.width=8, fig.height=5}
library(ggplot2)

ggplot(combined_gender_log, aes(x = RIDAGEYR, y = iron_log, col = Gender)) +
  geom_point(na.rm = TRUE, alpha = 0.4) +
  labs(
    x = "Age",
    y = "log iron intake",
    title = "Daily Average Supplemental Iron Intake by Age and Gender over 30 days"
  ) +
  theme_minimal() +
  geom_abline(intercept = 2.178715, slope = 0.001437, col = "blue") +
  geom_abline(intercept = 1.463298, slope = 0.007339, col = "red")
```

This demonstrates data transformation and visualization: recoding gender into a
categorical factor, transforming a bottom-dense variable with a natural
log, then using a scatter plot with fitted lines for each gender to explore the
association between two variables. Coloring by gender lets both
groups be compared in one figure, and `alpha = 0.4` makes overlapping regions
visible.

Overall the points show low iron intake across ages. The linear associations for
both genders are very weak, with slight slopes and low r-squared, indicating no
particular association between iron intake and age. Males show a slightly
steeper slope and higher r-squared, suggesting a marginally stronger
association.

## Plot 2: mean intake by 10-year age band

```{r binned-bar, fig.width=8, fig.height=5}
binned_data <- combined_gender_log %>%
  mutate(Age_Group = floor(RIDAGEYR / 10) * 10) %>%
  group_by(Gender, Age_Group) %>%
  summarise(mean_iron = mean(iron_log, na.rm = TRUE), .groups = "drop")

ggplot(binned_data, aes(x = factor(Age_Group), y = mean_iron, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Average Daily Iron Intake by Age Group and Gender",
    x = "Age Group",
    y = "Mean Iron Intake (log)",
    fill = "Gender"
  ) +
  theme_minimal()
```

Because the outcome is on a log scale, differences that look small are
substantial. The largest gap between genders appears early in life, with women
reporting much higher iron intake than men in the 10 and 20 year bins, which 
infer association with higher iron requirements during menstruation. Intake 
converges around the 50 year bin, when menopause occurs and less additional iron 
is needed. After that the pattern reverses, with men reporting higher intake in 
the 80 year bin, possibly reflecting anemia in elderly men.

## Statistical test

The outcome is continuous, highly right-skewed, and holds many zero values, and
the two groups (younger and older adults) are independent. A Wilcoxon rank-sum
test is therefore more appropriate than a two-sample t-test, which assumes
normality. The rank-sum test compares distributions using ranks, so a difference
between age groups can be assessed without a normal distribution.

**Null hypothesis.** The distribution of daily supplemental iron intake is the
same for older adults (age >= 50) and younger adults (age < 50).

**Alternative hypothesis.** The distribution differs between older and younger
adults.

```{r wilcoxon}
# prepare data, remove missing values, separate ages
iron_data <- combined_gender %>%
  filter(!is.na(DSQTIRON)) %>%
  mutate(Age_Group = ifelse(RIDAGEYR >= 50, "Older", "Younger"))

# split by gender so the tests run separately within each group
females <- iron_data %>% filter(Gender == "Female")
males   <- iron_data %>% filter(Gender == "Male")

wilcox_female <- wilcox.test(DSQTIRON ~ Age_Group, data = females)
wilcox_female

wilcox_male <- wilcox.test(DSQTIRON ~ Age_Group, data = males)
wilcox_male
```

```{r group-summary}
iron_data %>%
  group_by(Gender, Age_Group) %>%
  summarise(
    count  = n(),
    median = median(DSQTIRON, na.rm = TRUE),
    IQR    = IQR(DSQTIRON, na.rm = TRUE),
    .groups = "drop"
  )
```

## Plot 3: distribution by age group and gender

```{r violin, fig.width=8, fig.height=5}
ggplot(iron_data, aes(x = Age_Group, y = DSQTIRON, fill = Gender)) +
  geom_violin(trim = FALSE) +
  stat_summary(fun = median, geom = "point", shape = 21, size = 3,
               color = "white", fill = "red") +
  facet_wrap(~ Gender) +
  labs(
    title = "Daily Supplemental Iron Intake in mg by Age and Gender",
    subtitle = "Violin plot with red median points",
    x = "Age Group",
    y = "Daily Supplemental Iron Intake (mcg/dL)"
  ) +
  theme_minimal()
```

A violin plot presents the rank-sum results because it shows the full spread of
the data with the median marked in red. The male violins differ visibly from
each other, matching the statistically significant difference in iron intake
between older and younger men. The higher p-value for women indicates
insufficient evidence of a difference between younger and older women.

## Interpretation

For females the p-value was 0.2634, above 0.05, so there is not enough evidence
to reject the null hypothesis. For males the p-value was 0.01579, below 0.05, so
the null is rejected. Iron intake from supplements differs by age group among
men but not among women.

Several factors beyond age likely contribute. NHANES excludes people in
institutions such as hospitals, nursing homes, and prisons, which introduces
selection bias. The supplement data is self-reported and therefore subject to
inaccuracy. Confounding variables not included here, including health status,
income, and diet, could influence whether someone takes iron supplements and
could explain part of the observed difference.
